---
title: "Gastrulation: plot DNA methylation and chromatin accessibiliy dynamics associated with mesoderm commitment"
output: 
  BiocStyle::html_document: 
    fig_width: 10
    fig_height: 8
---

```{r}
library(data.table)
library(purrr)
library(ggplot2)
library(RColorBrewer)
library(splines)
```

```{r}
source("/Users/ricard/gastrulation/metaccrna/mesendoderm_commitment/mesoderm/load_settings.R")
source("/Users/ricard/gastrulation/metaccrna/mesendoderm_commitment/mesoderm/load_data.R")
```

<!-- Load results from differential methylation and accessibility analysis -->
```{r}
opts$diff.type <- 2
opts$min.fdr <- 0.10
opts$min.acc.diff <- 5
opts$min.met.diff <- 5

source("/Users/ricard/gastrulation/metaccrna/differential/load_data.R")
```

<!-- Subset differential sites from the data -->
Lineage-defining elements are defined as ChIP-seq peaks that show differential activity during germ layer commitment.
At 10% FDR, this represents around 25% of Endoderm and Mesoderm putative enhancers, as defined by the ChIP-seq.
Ectoderm enhancers show very few associations with germ layer commitment, so there is no point in subsetting them.

```{r}
# Methylation
met_dt <- met_dt %>% split(.$anno) %>%
  map2(.,names(.), function(x,y) x[id%in%diff.met[sig==T & anno==y,id]]) %>%
  rbindlist %>% droplevels()

# Accessibility
acc_dt <- acc_dt %>% split(.$anno) %>%
  map2(.,names(.), function(x,y) x[id%in%diff.acc[sig==T & anno==y,id]]) %>%
  rbindlist %>% droplevels()
```

<!-- Aggregate signal per cell (across all enhancers) -->
```{r}
met_dt <- met_dt[,.(rate=mean(rate)),by=c("id_met","anno","sample","stage","plate")]
acc_dt <- acc_dt[,.(rate=mean(rate)),by=c("id_acc","anno","sample","stage","plate")]
```

<!-- Regress out technical covariates -->

Methylation
```{r}
# Differences in global methylation rate
foo <- fread(io$met.stats) %>%
  .[,mean:=mean*100] %>%
  .[,c("id_met","mean")]
met_dt <- met_dt %>% merge(foo, by="id_met") %>%
  .[,rate:=mean(rate) + lm(formula=rate~mean)[["residuals"]], by=c("anno")] %>%
  .[,mean:=NULL]
```

Accessibility 
```{r}
# Differences in global accessibility rate
foo <- fread(io$acc.stats) %>%
  .[,c("id_acc","mean")]
acc_dt <- acc_dt %>% merge(foo, by="id_acc") %>%
  .[,rate:=mean(rate) + lm(formula=rate~mean)[["residuals"]], by=c("anno")] %>%
  .[,mean:=NULL]
```

<!-- Load precomputed pseudotime  -->
```{r}
Z <- fread(io$pseudotime) %>%
  .[,c("sample","x")] %>% setnames("sample","id_rna") %>%
  merge(sample_metadata[,c("sample","id_rna","lineage10x_2")], by="id_rna") %>% 
  setnames("x","Factor 1") %>% setorder("Factor 1") %>%
  .[,rank:=rev(1:.N)]
```

<!-- Rename annotations for plotting -->
```{r}
met_dt %>% .[,anno:=stringr::str_replace_all(anno,opts$views_names)] 
acc_dt %>% .[,anno:=stringr::str_replace_all(anno,opts$views_names)] 
```

<!-- Plot RNA pseudotime rank vs met/acc  -->

Both assays in one plot
```{r}
to.plot <- merge(met_dt, acc_dt, by=c("sample","anno","stage")) %>% 
  setnames(c("rate.x","rate.y"),c("met","acc")) %>%
  merge(Z[,c("sample","lineage10x_2","rank","Factor 1")], by="sample") %>%
  .[!(stage=="E7.5" & lineage10x_2=="Epiblast")] %>%
  melt(id.vars=c("sample","rank","anno","lineage10x_2","Factor 1"), measure.vars=c("met","acc"), variable.name="assay")
```

```{r}
p <- ggplot(to.plot, aes(x=`rank`, y=value)) +
  geom_point(aes(fill=assay), pch=21, color="black", alpha=0.9, size=1.0, stroke=0.05) +
  geom_rug(aes(color=lineage10x_2), sides="b") +
  facet_wrap(~anno, nrow=2, scales="free_y") +
  stat_smooth(aes(fill=assay), method="loess", color="black", alpha=0.75, span=0.3) +
  scale_fill_manual(values=c(acc="#00BFC4", met="#F37A71")) +
  scale_colour_manual(values=opts$colors) +
  # coord_cartesian(ylim=c(25,85)) +
  labs(x="Pseudotime (rank)", y="Methylation/Accessibility (%)") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.y = element_text(color="black", size=rel(1.0)),
    axis.text.x = element_blank(),
    axis.title = element_text(color="black", size=rel(1.2))
  )
# print(p)

# pdf(paste0(io$outdir,"/pseudotime_endoderm.pdf"), useDingbats = F, width=9, height=4)
p
# dev.off()
```

One plot per assay
```{r}
for (i in unique(to.plot$assay)) {
  
  p <- ggplot(to.plot[assay==i], aes(x=`rank`, y=value)) +
    geom_point(aes(fill=assay), pch=21, color="black", alpha=0.9, size=1.0, stroke=0.05) +
    facet_wrap(~anno, nrow=1, scales="fixed") +
    stat_smooth(aes(fill=assay), method="loess", color="black", alpha=0.75, span=0.5) +
    scale_fill_manual(values=c(acc="#00BFC4", met="#F37A71")) +
    scale_colour_manual(values=opts$colors) +
    labs(x="Pseudotime (rank)", y="Methylation/Accessibility (%)") +
    theme_classic() +
    theme(
      legend.position = "none",
      axis.text.y = element_text(color="black", size=rel(1.0)),
      axis.text.x = element_blank(),
      axis.title = element_text(color="black", size=rel(1.2))
    )
  
  if (i=="met") {
    p <- p +
      geom_segment(aes(x=680, xend=680, y=20, yend=100), size=0.25, color="black", linetype="dashed") +
      geom_segment(aes(x=900, xend=900, y=20, yend=100), size=0.25, color="black", linetype="dashed") +
      coord_cartesian(ylim=c(20,100))
  } else if (i=="acc") { 
    p <- p +
      geom_rug(aes(color=lineage10x_2), sides="b") + 
      geom_segment(aes(x=680, xend=680, y=25, yend=52), size=0.25, color="black", linetype="dashed") +
      geom_segment(aes(x=900, xend=900, y=25, yend=52), size=0.25, color="black", linetype="dashed") +
      coord_cartesian(ylim=c(25,52))
  }
  print(p)
  
  pdf(sprintf("%s/pseudotime_mesoderm_%s.pdf",io$outdir,i), useDingbats = F, width=9, height=2.5)
  print(p)
  dev.off()
}
```

Piece-wise linear regression
```{r}
knots <- c(680,900)
        
regression.out <- list()
for (i in unique(to.plot$anno)) {
  foo.met <- to.plot[anno==i & assay=="met"]
  foo.acc <- to.plot[anno==i & assay=="acc"]
  lm.met <- lm(formula = value ~ bs(rank, df=NULL, knots=knots, degree=1), data = foo.met)
  lm.acc <- lm(formula = value ~ bs(rank, df=NULL, knots=knots, degree=1), data = foo.acc)
  
  regression.out[[i]] <- data.table(
    anno = i, 
    knot = 1:(length(knots)+1), 
    met.pval = unname(summary(lm.met)$coefficients[-1,"Pr(>|t|)"]),
    met.r = unname(lm.met$coefficients[-1]),
    acc.pval = unname(summary(lm.acc)$coefficients[-1,"Pr(>|t|)"]),
    acc.r = unname(lm.acc$coefficients[-1])
  )
  
}
regression.out <- rbindlist(regression.out)
```

```{r}
foo <- regression.out %>% 
  melt(id.vars=c("anno","knot"), measure.vars=list(c("met.pval","acc.pval"),c("met.r","acc.r")), 
       variable.name="assay", value.name=c("pval","r"), variable.factor=F) %>%
  .[,assay:=ifelse(assay==1,"met","acc")] %>%
  .[,padj_fdr:= list(p.adjust(pval, method="bonferroni"))] 
```


