---
title: "Gastrulation scNMT: sex determination 1"
output: 
  BiocStyle::html_document:
  toc: false
fig_width: 10
fig_height: 8
---

Assign sex to embryos based on sum expression counts of genes on the Y chromosome.
This script only assigns sex to cells that can be traced back to individual embryos since the Y chromosome expression counts are not sufficiently informative for individual cells (e.g. not all male cells have Y chromosome counts).
To assign sex to cells that can not be traced back to individual embryos (mixed), run 'sex_determination_YRNA_Ymet.Rmd' after running this script.

```{r echo=FALSE, include=FALSE}
library(scater)
library(data.table)
library(purrr)
library(ggplot2)
```

```{r funcs, echo=FALSE}

```

```{r define_opts, echo=FALSE, include=FALSE}

## Define I/O ##
io <- list()
io$basedir <- "/Users/stapelc/Documents/gastrulation_data"
io$sample_metadata <- paste0(io$basedir,"/sample_metadata.txt")
io$gene.metadata <- "/Users/stapelc/Documents/GastrulaProject/data/ensembl/mouse/v93/BioMart/Mmusculus_genes_BioMart.93_GRCm38.p6.txt"
io$rna.infile <- paste(io$basedir,"rna/parsed/SingleCellExperiment.rds",sep="/")
io$outdir <- "/Users/stapelc/Documents/GastrulaProject/Results/Xchr"

## Define options ##
opts <- list()

# Define stages and lineages (only embryonic stages! X inactivation dynamics differ in extraembryonic cells (Xpi))
opts$stage <- c("E4.5","E5.5","E6.5","E7.5")
opts$lineage10x_2 <- c("Ectoderm","Embryonic_endoderm","Epiblast","Nascent_mesoderm","Mature_mesoderm","Primitive_Streak")

# opts$stage_lineage <- "all"
# if (opts$stage_lineage == "all") {
#   opts$stage_lineage <- fread(io$sample_metadata, header=T) %>% .[pass_rnaQC==T] %>% 
#     .[,stage_lineage:=paste(stage,lineage,sep="_")] %>% .$stage_lineage %>% unique
# }

# Define which cells to use
opts$cells <- fread(io$sample_metadata) %>% 
  .[,stage_lineage:=paste(stage,lineage10x_2,sep="_")] %>%
  .[pass_rnaQC==T & stage%in%opts$stage & lineage10x_2%in%opts$lineage10x_2,sample]

```

<!-- Load expression data -->
```{r load_expr, echo=FALSE, include=FALSE}

# Load SCEset object
sce <- readRDS(io$rna.infile)
colnames(sce) <- colData(sce)$sample

# Filter cells
sce <- sce[,colnames(sce) %in% opts$cells]

```

<!-- Load sample metadata -->
```{r}
sample_metadata <- fread(io$sample_metadata) %>% .[sample %in% opts$cells] %>% .[,stage_lineage:=paste(stage,lineage10x_2,sep="_")]
```

<!-- Extract gene metadata -->
```{r}
gene_metadata <- fread(io$gene.metadata, stringsAsFactors=T) #%>% 

#gene_metadata_1 <- ranges(sce) %>% as.data.table %>% setnames("names", "gene")
#gene_metadata_2 <- rowRanges(sce) %>% as.data.table %>% setnames(c("seqnames"), c("chr")) %>% 
#  .[,c("chr", "start", "end", "ens_id")] %>% .[,chr:=as.factor(sub("chr","",chr))]
#gene_metadata_all <- merge(gene_metadata_1, gene_metadata_2, by = c("start", "end"))
```

<!-- Create data.table from SCEset object -->
```{r}
rna <- counts(sce) %>% t %>% as.data.table(keep.rownames="sample") %>% melt(id.vars="sample", value.name="counts", variable.name="ens_id") #%>%
  #merge(fData(sce) %>% tibble::rownames_to_column("gene") %>% .[,c("gene","ens_id")] %>% as.data.table %>% setnames("ens_id","id"))
```

<!-- Merge data and metadata -->
```{r}
rna <- merge(rna, sample_metadata[,c("sample","stage_lineage","embryo")], by="sample") %>%
  merge(gene_metadata[,c("chr","ens_id","gene")], by="ens_id")
```

<!-- Filter expression data -->
```{r filter, echo=FALSE}

# Extract genes in the Y chromosome
rna_filt <- rna
rna_filt <- rna_filt[chr=="Y"]

```

<!-- plot expression of y-chr genes in different embyros per gene -->
```{r}
test <- merge(rna_filt, sex[,c("sample", "sex")], by="sample")

for (g in unique(rna_filt$gene)){
  p <- ggplot(rna_filt[gene==g], aes(x=embryo, y=counts)) +
    geom_point(pos="jitter") + 
    ggtitle(g) + 
    ylim(0,100) +
    theme(
    axis.title.y = element_text(colour="black", size=14, vjust=1.5),
    axis.title.x = element_text(colour="black", size=14, vjust=1.5),
    axis.text.x = element_text(angle=90, size=10, vjust=0.5, hjust=1.0, color=b$color),
    axis.text.y = element_text(colour="black",size=rel(1.3)),
    axis.line = element_line(colour="black", size=0.5),
    axis.ticks.x = element_line(colour="black", size=rel(0.8)),
    axis.ticks.y = element_line(colour="black", size=rel(0.8)),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )
  print(p)
}

y_genes <- c("Ddx3y","Eif2s3y","Kdm5d","Uty")

rna_filt <- rna_filt[gene %in% y_genes]

```

<!-- Determine sex of each cell and each embryo -->
```{r}
# define minimum number of mean Y-chr counts for male
opts$min.counts <- 1

# assign sex to individual cells
foo <- rna_filt[,.(counts=sum(counts)), by=c("sample" ,"embryo")]
females <- foo[counts<opts$min.counts]
foo$sex <- ifelse(foo$sample %in% females$sample, 0, 1)

# assign sex to individual embryos
bar <- foo[,.(sumSex=sum(sex), n=length(sex)), by="embryo"]
bar$ratio <- bar$sumSex / bar$n
bar$embryo_sex <- ifelse(bar$ratio < 0.2, "female", ifelse(bar$ratio > 0.8, "male", "unknown"))

# verify sex of individual cells based on sex assigned to the corresponding embryo
sex <- merge(foo, bar[, c("ratio", "embryo_sex", "embryo")], by = "embryo") 
sex$sex <- ifelse(sex$sex == 0, "female", "male")
sex$pass_sexQC <- ifelse(sex$sex == sex$embryo_sex, T, F)

```

<!-- Bar plot of number of counts in Y chr per embryo -->
```{r}

# mean Y counts per cell for each embryo
a <- sex[pass_sexQC == T]
b <- a[,.(counts=sum(counts), n=length(counts)), by="embryo"]
b$mean_counts <- b$counts/b$n

# sort embryos based on Y count levels
b <- b %>%  setkey(mean_counts) %>% .[,embryo:=factor(embryo,levels=embryo)]

# color bars based on sex
b$color <- c("blue","magenta")[as.numeric(b$mean_counts < opts$min.counts)+1]

# plot
p <- ggplot(b, aes(x=embryo, y=mean_counts)) + 
  ggtitle("") +
  geom_bar(stat='identity', position="dodge") +
  geom_hline(yintercept=opts$min.counts, colour="red", linetype="dashed") +
  xlab("Embryo") + ylab("Mean sum of counts \n per cell in Y chr") +
  theme(
    axis.title.y = element_text(colour="black", size=14, vjust=1.5),
    axis.title.x = element_text(colour="black", size=14, vjust=1.5),
    axis.text.x = element_text(angle=90, size=10, vjust=0.5, hjust=1.0, color=b$color),
    axis.text.y = element_text(colour="black",size=rel(1.3)),
    axis.line = element_line(colour="black", size=0.5),
    axis.ticks.x = element_line(colour="black", size=rel(0.8)),
    axis.ticks.y = element_line(colour="black", size=rel(0.8)),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )
print(p)

# save
#ggsave(filename=(paste0(io$outdir,"/scNMT_sex_countsYRNA",".pdf")), plot=p)


```

<!-- Assign sex to metadata -->
```{r}
# assign sex
sample_metadata_all <- fread(io$sample_metadata)
sample_metadata_sex <- merge(sample_metadata_all, sex[,c("sex", "pass_sexQC", "sample")], by="sample", all=TRUE)

# save new metadata including sex information
write.table(sample_metadata_sex,paste(io$outdir,"sample_metadata_sex.txt",sep="/"),sep="\t",row.names=FALSE)

```
